Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-17 versions/response.v1.txt

> Getting exact regions is what I think the authors should do. The
> approach used to obtain Fig 5 is generally applicable. Explicit,
> relatively simple, expressions are given in the appendix for long-run
> lnls, l_{tau_1,t*}(tau_2,w,x,y). For any given (x*,y*) on a fine grid
> in Figs 2-4, l_{tau_1,t*}(tau_j,w,x,y), can be maximized using
> numerical algorithms over the three parameters w,x,y. Doing so for
> tau_j=tau_1, tau_2 would determine the actual regions of problematic
> estimation corresponding to Figs 2-4. The only potential difficulty
> that I can see is that numerical algorithms might not converge. This
> can either be avoided by considering a cubic grid of starting (w,x,y),
> or through something like the following. For any given set of
> ancestral character state assignments for the distinct set of 8
> patterns 1111, 1110,... (0000 can treated as the same as 1111), the
> long-run lnl is of the form

> l(x,y,w)=a log(1-x) + b log(1+x) + c log(1-y) + d log(1+y) + e log(1-w) + f log(1+w)

> where a,..f>=0 are constants determined by (x*,y*) and what ancestral
> states have been assigned. The function l(x,y,w) has an explicit
> simple maximizer. Thus the overall maximizer is determined as the
> largest maximized l(x,y,w) over all 4^8 possible ancestral state
> assignments: a finite number of comparisons. The actual number of
> possible ancestral state assignments is quite a bit smaller than 4^8
> (for instance 11 is always the ancestral reconstruction for 1111) and
> the authors have a lot of familiarity with what the ancestral
> reconstructions are. I believe Fig 2 can be replaced with an exact (up
> to grid approximation) region of topological inconsistency and Figs
> 3-5 with heat maps of limiting (x,y,w) for tau_1.

Please see above for our response.
%das: I added some stuff below to maybe alleviate some confusion and to show why what this reviewer thinks we should do won't work, but not sure how much is necessary.
It is true that the form given above for l(x,y,w) has an explicit and simple maximizer.
However, the long-run log-likelihood is more complicated than this function, as seen in equation (9).
For both topologies, we have a number of equations of this form that we would need to maximize in a grid search.
The function of the form in l(x,y,w) above was our approximation to the upper and lower bounds of the log-likelihood, and was our choice for these bounds precisely because it has an explicit and simple maximizer.
Doing a grid search over all ancestral states would not be trivial, and we would need to perform numerical optimization for all possible ancestral states to compute the maximum.
Moreover, not assuming x_1=x_2 and y_1=y_2 (see below) results in an explosion of functions to perform a grid search over, with none of these functions being guaranteed concave.
The analytic bounds we have provided were devised specifically to sidestep this nonconcavity.

@@

> The other main concern that I have is with the restriction that
> t=(x,y,x,y,w) in estimation rather than t=(t_1,...,t_5). On the one
> hand, usually one expects that incorporating knowledge about the true
> underlying tree (t_1=t_3 & t_2=t_4 in the present case) should be
> helpful for a good estimation procedure. Thus the fact that treating
> ancestral states as parameters gives inconsistent topological
> estimation in this setting demonstrates that the approach is
> problematic. On the other hand, allowing t=(t_1,...,t_5) is what would
> be done in any practical implementation and it seems quite possible to
> me that results would be radically different in this case. For
> instance, from Fig 2, the generating tree with all equal edge-lengths
> (y*=x*=0.2) provides a case where topology tau_2 is inconsistently
> estimated with the restriction that estimated t=(x,y,x,y,w). Because
> of the symmetry of the problem, I am doubtful y*=x*=0.2 would be a
> point of inconsistency if t=(t_1,...,t_5) were unrestricted in
> estimation. The issue deserves discussion because I don't think it is
> transparent.

TODO: make the argument more clear in the main text.
%das: BIIIIG update
We thank this reviewer for highlighting an omission in our argument.
We had initially simplified the results by making the erroneous assumption that t_1=t_3 and t_2=t_4.
It can be shown that this assumption ignores potential feasible parameter sets, and we have removed it.
This requires a change in the arguments in the theorems as well, though the conclusions are similar.

@@

> I don't think the restricted case of p8-9 will be of interest to
> readers. Only the general case.

%EM I think we could move this to the appendix to throw this reviewer a bone.
%das: another big update
We have removed the restricted case as it does not reflect how inference is performed and was only included as an expository example.
Given the change to the theorem arguments, we do not believe this case is necessary for highlighting inconsistency.

@@

> Minor Comments:

> -l78-88: Calculations are more natural with `fidelities' and this
> might remain in the appendix but it is much more conventional in the
> literature for x,y,w to be edge-lengths or probabilities of a
> different character states at the tips of the edge. It takes quite a
> bit of getting used to realize that (0,0) corresponds to infinite
> edge-lengths and it seems unnecessary with a little work on the part
> of the authors to do the conversions that will make the figs more
> natural for readers.

Thank you for this suggestion.
We have implemented it in all the plots and have been careful to clarify that "fidelities" do not correspond to "lengths" in the text, deferring to "lengths" where appropriate.

@@

> Also, laying out the theory a little bit more step-by-step so that a
> general quantitative reader doesn't struggle too much to follow would
> help (suggestions below).

Thank you very much for reading the derivations and for your suggestions. As noted inline we have revised the text accordingly.

> I know that it's not feasible to explicitly compute (analytically)
> these likelihoods for bigger trees. But one of the main motivations
> and applications for tools like TreeTime is surely that the number of
> sequences grows (not the number of sites, which is fixed by the viral
> sequence length in the case of viruses). What happens when the number
> of tips grows? Are there more opportunities for the kind of issue that
> gives the wrong value for w? Many influenza trees have many 0 (or near
> 0) internal branch lengths, and it would be interesting to understand
> the implications for the ancestral state reconstructions (which in
> turn are used to infer spatial spread, transmission etc).

%EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?
%das: I wasn't able to come up with anything interesting for this.

@@

> L 132 clarify “equivalent but different notion” – different from what?

We have reformulated this sentence to avoid the comparison.

> - line 146 “seeing no change along a branch more likely increases the
> likelihood by introducing a term of (1 + θ) instead of (1 − θ)” – I
> don't understand this

TODO: this is changed, but I don't see this being a lot clearer.
%das: I believe this is clear now; lmk
We have clarified the intuition in this sentence.

@@

> line 364 define p(e)

Thanks for reminding us to define p(e), which is now described as "where $p(e)$ is the probability of a character change along edge $e$."

> line 366  I don't understand (8) but perhaps that's just background I
> don't have – a quick explanation or reference would be handy
> What's a 'generating probability' ? Is that the fidelity? Or a
> generator?

We have updated the text to explicitly define "generating probability." Background for equation (8) can be found in Section 8.6 in the Semple and Steel text. This is referenced at the beginning of the section. We do not believe an exhaustive explanation would provide any benefit to the reader that they could not get from referring to the text.

@@

> Reviewer: 3

> Comments to the Author
> Shaw and Matsen show that a joint maximum-likelihood inference of
> phylogenies and ancestral states is not consistent when branch lengths
> are long. We have recently proposed such a joint estimation for
> sequences that recently diverged and are hence similar (Sagulenko et
> al). The results presented here are anticipated in Sagulenko et al
> "For more divergent sequences, however, subleading states of internal
> nodes make a substantial contribution and the iterative optimization
> will underestimate the branch lengths."
> and
> "The iterative optimization procedures are not expected to be accurate
> for trees were many sites are saturated. In scenarios with extensive
> uncertainty of ancestral states and tree topology, convergence of the
> iterative steps cannot be guaranteed."

> In this light, it should be pointed out that joint estimation
> procedure was explicitly suggested as an approximation and hence never
> expected to be consistent in a strict sense. Furthermore, it was
> suggested as an approximation when branch lengths are short -- the
> opposite of the problematic regions identified here.
> But systematic analysis of the behavior of this approximation is
> useful and I believe the work is correct (but haven't checked the
> proofs). The work by Shaw and Matsen suggests the radius of
> convergence of this procedure is somewhere around branch length of
> order 10-20%, which is consistent with results reported by Sagulenko
> et al.

Thank you for your thoughts, and we have worked hard to not oversell our results.
The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
However, in our assessment the word "approximate" doesn't completely describe a situation in which there is significant bias.
The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but even in these cases the bias in branch lengths persists.

TODO we should make some changes to at least say that we've further clarified the implications of our results.
%das: how much more should we say than this sentence (which we already say)? We already quote them as saying they should only use it when branches are short.
%Although the software described in \citet{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
%and
%In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
%This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.

@@

> At the same time, these results are not particularly surprising.
> Assigning ML ancestral states will always result in shorter branches.
> For short branches, this correction is linear in branch length, for
> longer branches higher order effects start to contribute. If the
> typical branch length around a bifurcating node is x<<1, a fraction ~x
> of the sites mutated along one of the three adjacent branches, while a
> fraction ~x^2 mutated on two branches and not the third. The latter
> case is ignored in this joint inference procedure. Not having to sum
> over these subleading states is what makes joint inference faster, but
> such truncation will first lead to inaccuracies and at longer
> distances to qualitatively different solutions (like Taylor expanding
> 1/(1-x) ).
>
> There are other obvious cases where joint inference will fail. For
> sequences with a heavily biased in nucleotide composition (say biased
> towards A). For sufficiently long terminal branches, all ancestral
> states will be estimated as all 'A' and subsequent branch length
> estimate will shrink all internal branches to zero. This happens for
> the very same reason that the most likely outcome of multiple flips of
> a biased coin is all heads -- a very atypical yet most likely outcome.
>
> Currently, the work largely shows that areas of inconsistencies exist.
> It would be useful if the authors could present Taylor expansions for
> the ML branch length estimates around x=y=w=1. This would more
> explicitly show the inaccuracies accumulate with increasing branch
> length and how the inaccuracies depend on asymmetries in the tree. To
> me, this would be much more useful than to know that joint inference
> will fail if branches are long.

This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (Lemma 3 in the updated paper).
We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree.
Asymmetries can drive these particular choices, but Fig. 3 in the revised manuscript shows that almost half of the parameter space results in branch lengths being estimated as zero.
We do not think providing an argument about accumulating inaccuracies would elucidate why such starkly-defined regions of inconsistency exist.
For reasonable parameter settings, one could investigate the error accumulation given fixed ancestral states, but this was not our goal and will require another paper.

TODO describe why Taylor isn't a straightforward thing.
TODO refer to idea of applied paper again?

@@

> Specific comments:
>  - when y=x=w, the two topologies are indistinguishable other than for
>  the labeling of the leaves. What happens in Fig 2 on the diagonal?
>  Does it degenerate into a star?

It is true when y=x=w we are unable to distinguish the topologies apart from the labeling, though the labeling is what distinguishes the topologies.
As shown in Goldman (1990), with all branch lengths fixed and equal the joint inference procedure degenerates into parsimony, which has well known inconsistency issues.
We have added discussion of this point.

@@

>  - line 50: overestimates branch lengths? It overestimates fidelity ->
>  underestimates branch length.  As expected.  Again line 249. Please
>  check. The back and forth between fidelity and branch length can be
>  confusing.

Thank you for bringing this to our attention.
We have tried to use branch lengths in as many instances as we could in the discussions, using fidelities only in the computations as they allow for great simplifications.

@@

>  - Fig 5: it is surprising that \hat{w} seems completely flat as a
>  function of x* until it jumps discontinuously to 0. Why not include
>  the analytical bound in Fig 5?

The discontinuity is a reflection of the change in ancestral states when the generating parameters change.
As the true generating parameters change, we become more likely to choose a different ancestral state as most likely, resulting in an entirely different objective function to be optimized.
The analytical bound is a coarse estimate of where inconsistency arises.
In the revised manuscript, we fix the maximal ancestral state to simplify the argument, and comparing the analytical with the empirical would not provide much added benefit to the analysis.

@@

>  - It would be nice if the abstract pointed out that these
>  inconsistencies arise when branch lengths are long and that the
>  approximate joint maximization procedure was suggested for precisely
>  the opposite limit. Similar at the beginning of the discussion, I
>  think it would be more appropriate if this would say smth along the
>  lines "Sagulenko et al suggested an approximation for short
>  branches... we show that this doesn't work for long branches..."

TODO we should do some change, but make it clear that it's consistently biased even for short branches.
%das: I added this line:
We have added the line:
\citet{Sagulenko2017-jo} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.

We believe it's important not to mischaracterize the suggested use of joint inference, though we would like to emphasize to the reader that problems are apparent in large regions of the parameter space.
We think the current discussion treats this optimization procedure as fairly as possible.

@@

Thank you for your assessment.
We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and we would appreciate it if you could take a look at the revision.

> The only real criticism I have of the paper is that the authors could
> have done a better job in connecting their new work to previous
> results.  For example, Goldman (1990, Syst. Biol. 39:345-361) showed
> that joint estimation of topology and ancestral states by maximum
> likelihood was equivalent to maximum parsimony analysis for the same
> 2-state model considered by Shaw and Matsen when all branches are
> forced to have the same length.  They noted the statistical
> inconsistency of this approach, which was necessarily true because
> parsimony was already known to be inconsistent (the famous result of
> Felsenstein 1978).  Shaw and Matsen extend this result to the case of
> unequal branch lengths, which while not particularly surprising, is
> nonetheless an important extension.  The authors should also cite key
> relevant earlier work on the identifiability ot the tree topology with
> "classical" maximum likelihood phylogenetic estimation (Allman et al
> 2008, Adv Appl Prob 40:229–249; Chai and Housworth 2011, Syst Biol
> 60:713–718) in addition to crediting Roychoudhury 2015 for
> establishing ML consistency.

We are very grateful to this reviewer for pointing out the Goldman paper.
We were not aware of this paper and now have a paragraph describing it and comparing to our work.
It's also a good idea to cite additional papers about identifiability, and we have done so.

%EM David, heh, can you point me to the place where the equal branch length result was done?
%EM Also, I think you said that he says that the varying branch length thing is an open question-- it would be great to quote that.
%das: update I added some of the below. Let me know what you think. And re: Allman/Chai, I just added them in with RoyChoudhury; is that ok? They are all just examples of conditions where phylogenetics is consistent it seems.

%das: below are some notes and quotes. I will work on paring these down and adding some to the text, maybe with a little more diplomacy.
%The equal branch length example occurs in Example 2, starting halfway down the second column of page 352 in Goldman (1990) and continuing until the top of the second column on page 354.
%Here he only talks about how parsimony is inconsistent because it may result in equally parsimonious trees (this is the last paragraph of the example section).
%It's not until the bottom of page 356 that he punts the inconsistency to Felsenstein and Hendy/Penny, and the bottom of page 357 where he wonders how small branch lengths have to be for parsimony to be consistent in general.

### 2018-07-17 joint_inf.tex
consistent notation and a few loose ends
pass thru reviewer response

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral state splits.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of the unknown branch parameters $t$.
This tendency to include $(1+w)$ terms in the likelihood results in a bias toward one of branch fidelities.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology, as the true $y^*\neq 1$.

@@

As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
To understand the quality of each fit, we report the range of $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference, $[-3\times 10^{-17}, 3\times 10^{-17}]$ showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

>>>>

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral state splits.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
As $(1+w) > (1-w)$, a maximization procedure more likely introduces the $(1+w)$ term, though this will not be guaranteed as the maximum depends on the values of the unknown branch parameters $t$.
This tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology, as the true $y^*\neq 1$.

@@

As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
%TODO: does this bias make sense? should it be log-diff?
To understand the quality of each fit, we report the range of $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference, $[-3\times 10^{-17}, 3\times 10^{-17}]$ showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
